A general Nullstellensatz for generalised spaces

### 2019-07-07 slides-edinburgh2019.tex
Slides for CT 2019 (almost finished)

\usepackage{pgfpages}
\setbeameroption{show notes on second screen}
\setbeamertemplate{note page}[plain]
\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}

@@

\setbeamertemplate{frametitle}{%
  \vskip0.4em%
  \leavevmode%
  \begin{beamercolorbox}[dp=1ex,center]{}%
  %   \usebeamercolor[fg]{item}{\textbf{{\Large \insertframetitle}}}
    \begin{tikzpicture}
      \def\R{8pt}
      \node (title) {\hil{\large\insertframetitle}};
      \begin{pgfonlayer}{background}
        \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
          ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
          ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
          ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
          ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
          cycle;
      \end{pgfonlayer}
    \end{tikzpicture}
  \end{beamercolorbox}%
  \vskip-0.6em%
}

@@

  \bigskip
  \includegraphics[width=0.4\textwidth]{phantoms}
  \bigskip

  \begin{tikzpicture}
    \def\R{8pt}
    \node (title) {\hil{A general Nullstellensatz for generalised spaces}};
    \begin{pgfonlayer}{background}
      \draw[decoration={bumps,segment length=8pt}, decorate, very thick, draw=mypurple]
        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
        cycle;
    \end{pgfonlayer}
  \end{tikzpicture}

@@

  6th Workshop on Formal Topology in Birmingham \\
  April 11th, 2019
  \bigskip

@@

  \begin{quote}
    \emph{\mbox{For any finitely presented~$\affl$-algebra~$A$, the canonical map}
    \[ A \longrightarrow (\affl)^{\Hom_\affl(A,\affl)},\ s \longmapsto (x
    \mapsto x(s)) \]
    is an isomorphism of~$\affl$-algebras.}
  \end{quote}

@@

    The asterisks in \emph{geometric$^\star$ sequent} and
    \emph{provability$^\star$} indicate that any infinities used to index
    disjunctions have to be come from the base topos. This restriction is an important
    subtlety, though not vital to this talk. If~$\TT$ is a coherent
    theory, then for coherent sequents there is no difference between
    provability in coherent logic, provability in geometric logic and
    provability$^\star$.
  }

@@

\begin{frame}{Revisiting the test cases}
  \nullstellensatz
  \medskip

  \textbf{In the object classifier.}
%  Assume that~$U_\TT$ is empty, that is $\forall x \? U_\TT\_ (\top \Rightarrow
%  \bot)$. By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$(\top \seq{x} \bot)$.
%  But this is false in the~$\ul{\TT}/U_\TT$-model~$U_\TT \amalg
%  \{\heartsuit\}$.
  Let~$x,y \? U_\TT$. Assume that~$\neg(x = y)$.
  By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$(e_x = e_y \vdash \bot)$.
  But this is false in the~$\ul{\TT}/U_\TT$-model~$U_\TT/(x \sim y)$.
  \medskip

  \textbf{In the ring classifier.}
  Let~$f,g \? U_\TT[X]$ such that any zero of~$f$ is a zero of~$g$.
  By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves this fact.
  Hence it holds in the~$\ul{\TT}/U_\TT$-model~$U_\TT[X]/(f)$. In this
  model~$f$ has the zero~$[X]$. Hence also~$g([X]) = 0$
  in~$U_\TT[X]/(f)$, that is~$g = hf$ for some~$h \? U_\TT[X]$.

@@

    Since the converse direction in the Nullstellensatz also holds
    (because~$U_\TT$ is a~$\ul{\TT}/U_\TT$-model), the
    statements~$\speak{$\sigma$ holds for~$U_\TT$}$
    and~$\speak{$\ul{\TT}/U_\TT$ proves$^\star$ $\sigma$}$ are equivalent. This
    equivalence is intriguing from a logical point of view, since the former
    statement is a geometric implication while the latter can be put as a
    geometric formula. (Up to a subtle issue indicated on the next slide.)

@@

    Often when using the Nullstellensatz, we go from an (assumed) truth
    of~$U_\TT$ via provability$^\star$ to another model~$M$
    of~$\ul{\TT}/U_\TT$. That is, we use provability$^\star$ as a (one-way)
    \emph{bridge}:
    \[ \xymatrix{
      & \text{$\ul{\TT}/U_\TT$ proves$^\star$ $\sigma$} \\
      \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
      \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
    } \]
  }
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\hfill\mbox{\includegraphics[width=0.13\paperwidth]{hen}\;\;}\end{minipage}}
\begin{frame}{Exhaustion and extensions}
  \justifying
  \textbf{Theorem 1.} A first-order formula holds for~$U_\TT$ iff it is
  intuitionistically provable from the axioms of~$\TT$ and
  the scheme
  \begin{equation}\tag{Nullstellensatz} \speak{$\sigma$ holds} \implies \speak{$\ul{\TT}/U_\TT$
  proves$^{\star\star}$~$\sigma$}. \end{equation}

  \textbf{Theorem 2.}
  Let~$\TT'$ be a quotient theory
  of~$\TT$. Assume that~$U_\TT$ is contained in the subtopos~$\Set[\TT']$. Then internally
  to~$\Set[\TT']$:

  \vspace*{-0.4em}
  \begin{quote}
    \emph{A geometric$^\star$ sequent~$\sigma$ with Horn consequent holds for~$U_{\TT'}$
    iff~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.}
  \end{quote}

  \textbf{Theorem 3.}
  % Let~$\mathrm{Form}^\star_{x\?X}(\ul{\TT}/U_\TT)/(\dashv\vdash_{x\?X})$ be the~$\Set[\TT]$-object of
  % geometric$^\star$ formulas over the signature of~$\ul{\TT}/U_\TT$ in the
  % context~$x\?X$, where any two such formulas are identified if and only
  % if~$\ul{\TT}/U_\TT$ proves them equivalent.
  The morphism~$\mathrm{ev}$ below is an isomorphism.
  \begin{equation}\label{eq:hon}\tag{$\star$}
    \mathrm{ev} : \mathrm{FunctFormulas}^\star(\ul{\TT}/U_\TT)/(\dashv\vdash) \longrightarrow P(U_\TT)
  \end{equation}

  \textbf{Theorem 4.} A higher-order formula holds for~$U_\TT$ iff it is
  provable in intuitionistic higher-order logic from the axioms of~$\TT$ and
  the higher-order Nullstellensatz scheme~($\star$).

>>>>

%\usepackage{pgfpages}
%\setbeameroption{show notes on second screen}
%\setbeamertemplate{note page}[plain]
\newcommand{\jnote}[2]{}

@@

\setbeamertemplate{frametitle}{%
  \vskip0.4em%
  \leavevmode%
  \begin{beamercolorbox}[dp=1ex,center]{}%
  %   \usebeamercolor[fg]{item}{\textbf{{\Large \insertframetitle}}}
    \begin{tikzpicture}
      \def\R{8pt}
      \node (title) {\hil{\large\insertframetitle}};
      \begin{pgfonlayer}{background}
        \draw[decorate, very thick, draw=mypurple]
          ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
          ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
          ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
          ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
          cycle;
      \end{pgfonlayer}
    \end{tikzpicture}
  \end{beamercolorbox}%
  \vskip-0.6em%
}

@@

  \bigskip
  \[ \xymatrix{
    & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
    \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
    \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
  } \]
  \bigskip

  \begin{tikzpicture}
    \def\R{8pt}
    \node (title) {\hil{A general Nullstellensatz for generalised spaces}};
    \begin{pgfonlayer}{background}
      \draw[decorate, very thick, draw=mypurple]
        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
        cycle;
    \end{pgfonlayer}
  \end{tikzpicture}

@@

  Category Theory 2019 in Edinburgh \\
  July 9th, 2019
  \bigskip

@@

  \begin{quote}
    \emph{\mbox{For any finitely presented~$\affl$-algebra~$A$, the ``double dual'' map}
    \[ A \longrightarrow (\affl)^{\Hom_\affl(A,\affl)},\ s \longmapsto (x
    \mapsto x(s)) \]
    is an isomorphism of~$\affl$-algebras.}
  \end{quote}

@@

    The asterisks in \emph{geometric$^\star$ sequent}
    indicate that any infinities used to index
    disjunctions have to be come from the base topos. This restriction is an important
    subtlety, though not vital to this talk. If~$\TT$ is a coherent
    theory, then for coherent sequents there is no difference between
    provability in coherent logic, provability in geometric logic and
    provability$^\star$.
  }

@@

\begin{frame}{Revisiting the test cases}
  \nullstellensatz
  \[ \xymatrix{
    & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
    \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
    \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
  } \]

  \only<1>{
    \textbf{In the object classifier.}
  %  Assume that~$U_\TT$ is empty, that is $\forall x \? U_\TT\_ (\top \Rightarrow
  %  \bot)$. By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$(\top \seq{x} \bot)$.
  %  But this is false in the~$\ul{\TT}/U_\TT$-model~$U_\TT \amalg
  %  \{\heartsuit\}$.
    Let~$x,y \? U_\TT$. Assume that~$\neg(x = y)$.
    By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves~$\sigma \defeq (e_x = e_y \vdash \bot)$.
    But this is false in the~$\ul{\TT}/U_\TT$-model~$M \defeq U_\TT/(x \sim y)$.
  }

  \only<2->{
    \textbf{In the ring classifier.}
    Let~$f,g \? U_\TT[X]$ such that any zero of~$f$ is a zero of~$g$.
    By the Nullstellensatz~$\ul{\TT}/U_\TT$ proves this fact.
    Hence it holds in the~$\ul{\TT}/U_\TT$-model~$U_\TT[X]/(f)$. In this
    model~$f$ has the zero~$[X]$. Hence also~$g([X]) = 0$
    in~$U_\TT[X]/(f)$, that is~$g = hf$ for some~$h \? U_\TT[X]$.
  }

@@

    Since the converse direction in the Nullstellensatz also holds
    (because~$U_\TT$ is a~$\ul{\TT}/U_\TT$-model), the
    statements~$\speak{$\sigma$ holds for~$U_\TT$}$
    and~$\speak{$\ul{\TT}/U_\TT$ proves $\sigma$}$ are equivalent. This
    equivalence is intriguing from a logical point of view, since the former
    statement is a geometric implication while the latter can be put as a
    geometric formula. (Up to a subtle issue indicated on the next slide.)

@@

    Often when using the Nullstellensatz, we go from an (assumed) truth
    of~$U_\TT$ via provability to another model~$M$
    of~$\ul{\TT}/U_\TT$. That is, we use provability$^\star$ as a (one-way)
    \emph{bridge}:
    \[ \xymatrix{
      & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
      \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
      \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
    } \]
  }
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{7.95cm}\hfill\mbox{\includegraphics[width=0.13\paperwidth]{hen}\;\;}\end{minipage}}
\begin{frame}{Exhaustion and extensions}
  \justifying
  \textbf{Theorem 1.} A first-order formula holds for~$U_\TT$ iff it is
  intuitionistically provable from the axioms of~$\TT$ and
  the scheme
  \begin{equation}\tag{Nullstellensatz} \speak{$\sigma$ holds} \implies \speak{$\ul{\TT}/U_\TT$
  proves~$\sigma$}. \end{equation}
  \vspace*{-1.5em}

  \textbf{Theorem 2.}
  Let~$\TT'$ be a quotient theory
  of~$\TT$. Assume that~$U_\TT$ is contained in the subtopos~$\Set[\TT']$. Then internally
  to~$\Set[\TT']$:
  \begin{quote}
    \emph{A geometric$^\star$ sequent~$\sigma$ with Horn consequent holds for~$U_{\TT'}$
    iff~$\ul{\TT}/U_\TT$ proves~$\sigma$.}\par
  \end{quote}

  \textbf{Theorem 3.} A higher-order formula holds for~$U_\TT$ iff it is
  provable in intuitionistic higher-order logic from the axioms of~$\TT$ and
  the higher-order Nullstellensatz scheme.
